---
title: "upscaler"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{upscaler}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(upscaler)
```

#### Introduction to `upscaler`
The `upscaler` package grew out of the need to teach Computational Biology students how to move beyond "single script" programming and build larger programs that operate on multiple files and call on separate scripts. No matter how large or small the project, code should be organized to create a reproducible research pipeline, so that all of the numerical input and output for a project can be reproduced, and the work can be reliably repeated if there are changes to the data or code. `upscaler` follows the general principles of functional programming, providing a small collection of R functions to reduce overhead and more easily document code and procedures. 

#### Acknowledgements
Package development was supported by NSF grant 2019470 RII Track-2 FEC : Harnessing Spatiotemporal Data Science to Predict Responses of Biodiversity and Rural Communities under Climate Change (co-PIs Brian J. McGill and Nicholas J. Gotelli). The functions were first presented at a Data Science Skill-building Bootcamp offered through the grant in June of 2022. I thank Alex Burnham, Leo Edminster-Cyr, Laurent Hebert-Dufrense, Matt Dube, Matt Kling, Bailey McLaughlin, and Brian McGill and the entire `barracuda` team (**B**iodiversity **a**nd **R**ural **R**esponse to **C**limate **C**hange **U**sing **D**ata **A**nalysis) for advice and improvements to the code.
